Mining the human microbiome for clinical insight

### 2018-12-23 aspiration/plosone_aspiration_main.tex
Mostly cosmetic changes, mostly singlespace bibs
Small fixes

Previous studies have shown that the mouth, upper respiratory tract, and lung microbiota contain similar microbes, and that upstream oral communities seed downstream sites (e.g. lungs and stomach) \cite{Bassis2015source,Charlson2011topographical,rosen2015ppi}.
However, there is little consensus on whether there exists a distinct or "core" lung microbiome that is consistent across people \cite{Charlson2011topographical,venkataraman-2015-dispersal,segal-2013-pneumotypes,erbDownward-2011-COPD}.
Most studies, however, agree that the lung microbial communities share taxa with the oral microbiome, but that there are some bacteria present in lung communities whose abundances cannot be traced solely to the mouth \cite{Bassis2015source,Charlson2011topographical,morris-2013-healthsmokers,segal-2013-pneumotypes}.

@@

\bibliographystyle{unsrtnat}
\bibliography{aspiration/aspiration_refs.bib}

%\end{document}

>>>>

Previous studies have shown that the mouth, upper respiratory tract, and lung microbiota contain similar microbes, and that upstream oral communities seed downstream sites (e.g. lungs and stomach) \cite{Bassis2015source,Charlson2011topographical,rosen2015ppi}.
However, there is little consensus on whether there exists a distinct or ``core'' lung microbiome that is consistent across people \cite{Charlson2011topographical,venkataraman-2015-dispersal,segal-2013-pneumotypes,erbDownward-2011-COPD}.
Most studies, however, agree that the lung microbial communities share taxa with the oral microbiome, but that there are some bacteria present in lung communities whose abundances cannot be traced solely to the mouth \cite{Bassis2015source,Charlson2011topographical,morris-2013-healthsmokers,segal-2013-pneumotypes}.

@@

\begin{singlespace}
\bibliographystyle{unsrtnat}
\bibliography{aspiration/aspiration_refs.bib}
\end{singlespace}

### 2018-12-23 chap_24hr.tex
Add percnorm and 24hr to intro, label all chapters
First draft of 24hr (not compiling yet)
Working version of 24hr chapter and supplement
Small fixes

%\graphicspath{{donor-selection/figures/}}

\chapter{Residential sewage provides a view into community-level human microbiome and metabolome}

@@

\bigskip
\bigskip
\noindent
The contents of this chapter constitute a manuscript in preparation for submission.

\clearpage
%\input{donor-selection/ms.tex}

>>>>

% This graphicspath isn't working for some reason,
% need to manually write paths for each figure
\graphicspath{{24hr/figures/}}

\chapter{Residential sewage provides a view into community-level human microbiome and metabolome}\label{chap:24hr}

@@

\bigskip
\bigskip
\noindent
The contents of this chapter constitute a manuscript in preparation for submission. The main figures are at the end of the chapter and the supplementary figures are in Appendix \ref{app:24hr}.

\clearpage
\input{24hr/24hr_ms.tex}

### 2018-12-23 chap_donor-selection.tex
Prep to add 24hr and perc norm chapters
Add percnorm and 24hr to intro, label all chapters
Small fixes

\chapter{Increasing fecal microbiota transplant clinical trial successes by applying rational donor selection}

@@

\bigskip
\noindent
The contents of this chapter constitute a manuscript in preparation for submission.

\bigskip
\input{donor-selection/ms.tex}

>>>>

\chapter{Increasing fecal microbiota transplant clinical trial successes by applying rational donor selection}\label{chap:donor-selection}

@@

\bigskip
\bigskip
\noindent
The contents of this chapter constitute a manuscript in preparation for submission.

\bigskip
\bigskip
\noindent
The figures appear within the main text of the chapter, and supplementary figures are at the end of the chapter.

\clearpage
\input{donor-selection/ms.tex}

### 2018-12-23 donor-selection/ms.tex
Small fixes

Fecal Microbiota Transplantation (FMT) is the transfer of healthy gut bacteria through whole stool into an individual with a diseased microbiome.
Early clinical successes combined with emerging research linking the microbiome to many different diseases are driving enthusiasm for FMT as a clinical treatment for complex conditions.
However, preliminary trials in exploratory indications have yielded mixed results and suggest that heterogeneity in donor stool may play a role in patient response.
Thus, as FMT expands to complex diseases, clinical trials with randomly-selected donors may fail because an ineffective donor was chosen rather than because FMT is not appropriate for the indication.
Here, we describe a conceptual framework to guide rational donor selection to increase the likelihood that FMT clinical trials will succeed.
We argue that the mechanism by which the microbiome is hypothesized to be associated with a given indication should inform how donors are selected for FMT trials.
We describe different disease models which may underlie microbiome-mediated conditions and associated rational donor selection strategies for each, and also provide examples based on previously published FMT trials and ongoing investigations.
Finally, we show that performing discovery-based retrospective research after an FMT clinical trial concludes may require prohibitively large sample sizes, suggesting that clinicians should focus on hypothesis-driven retrospective analyses and plan for these during the clinical trial design process.
In this nascent field, as FMT trials enter new disease indications, a rational approach to donor selection is urgently needed to maximize the insights generated to benefit patients.

@@

In some conditions, there may not be enough understanding of the underlying microbiome-based etiology to inform donor selection in an FMT trial. It may also be the case that there are no existing datasets on which to train models, existing datasets are not sufficiently powered to distinguish the different potential underlying models, or logistical considerations constrain a clinician's ability to select specific donors for their trial. In these cases, we recommend selecting different healthy donors, employing an adaptive clinical trial design in which donors are cycled after they have clinical failures (as described previously in Olesen, Gurry, and Alm 2017), and performing retrospective analyses to answer targeted hypotheses which were developed during the clinical trial design process.

@@

As donors change through the course of an adaptive trial, clinicians may elect to select their donors randomly or to more rationally cycle through donors (Olesen, Gurry, Alm 2017). ``Differently healthy'' donors may be selected, perhaps representing different underlying disease-associated models described above. Donors may also be selected to span the range of ``healthy'' microbiomes in a given population. For example, clinicians may pick a ``median'' health donor (i.e. one who is about as similar to all reference microbiomes), define a ``healthy plane'' and pick donors based on their distance to this plane (as in Halfvarson et al 2017), or simply based on the presence or abundance of certain consistently ``core'' health-associated bacteria (Shade and Handelsman 2012; Duvallet et al. 2017). In a similar vein, ``healthy'' donors can also be chosen based on their distance from disease-associated microbiomes, as opposed or in addition to similarity to health. Published case-control datasets can be used to identify donors with communities which are farthest away from the median or average diseased patient. These datasets can also be mined to identify taxa which are consistently disease-associated, and which should be minimized or perhaps even absent in the potential donor. Pairing rational donor selection with adaptive trial designs may eventually yield insight into the underlying model mediating the disease of interest if certain types of ``healthy'' donors consistently perform better at treating patients than others.

@@

The framework presented here encourages clinicians to leverage their clinical experience, existing microbiome research and published datasets, and the increasing availability of screened donor stools to more efficiently translate microbiome-based interventions into clinical impact.
Clinicians can apply their existing knowledge and a priori hypotheses to determine which microbiome-mediated disease model may underlie their indication of interest, and then select donors accordingly.
By rationally choosing donors during the FMT trial design, clinicians will increase the likelihood of successful FMT trials in diseases in which donor heterogeneity affects patient response.
Our power simulation analysis also suggests that retrospective analyses of the microbiome data generated should be developed during trial design, with targeted hypotheses of interest and sample collection plans tailored accordingly.
Otherwise, untargeted discovery-based analyses are unlikely to yield key information from most FMT trials.
Paired with adaptive clinical trial designs, FMT trials with rationally-selected donors will become an important tool in advancing translational microbiome research and clinical treatment to improve and save patient lives.

>>>>

Early clinical successes are driving enthusiasm for fecal microbiota transplantation (FMT), the transfer of healthy gut bacteria through whole stool, as emerging research is linking the microbiome to many different diseases.
However, preliminary trials have yielded mixed results and suggest that heterogeneity in donor stool may play a role in patient response.
Thus, clinical trials may fail because an ineffective donor was chosen rather than because FMT is not appropriate for the indication.
Here, we describe a conceptual framework to guide rational donor selection to increase the likelihood that FMT clinical trials will succeed.
We argue that the mechanism by which the microbiome is hypothesized to be associated with a given indication should inform how donors are selected for FMT trials, categorizing these mechanisms into four disease models and presenting associated donor selection strategies.
We next walk through examples based on previously published FMT trials and ongoing investigations to illustrate how donor selection might occur in practice.
Finally, we show that typical FMT trials are not powered to discover individual taxa mediating patient responses, suggesting that clinicians should develop targeted hypotheses for retrospective analyses and design their clinical trials accordingly.
Moving forward, developing and applying novel clinical trial design methodologies like rational donor selection will be necessary to ensure that FMT successfully translates into clinical impact.

@@

In some conditions, there may not be enough understanding of the underlying microbiome-based etiology to inform donor selection in an FMT trial. It may also be the case that there are no existing datasets on which to train models, existing datasets are not sufficiently powered to distinguish the different potential underlying models, or logistical considerations constrain a clinician's ability to select specific donors for their trial. In these cases, we recommend selecting different healthy donors, employing an adaptive clinical trial design in which donors are cycled after they have clinical failures (as described previously in \cite{Olesen2017}), and performing retrospective analyses to answer targeted hypotheses which were developed during the clinical trial design process.

@@

As donors change through the course of an adaptive trial, clinicians may elect to select their donors randomly or to more rationally cycle through donors \cite{Olesen2017}.
``Differently healthy'' donors may be selected, perhaps representing different underlying disease-associated models described above.
Donors may also be selected to span the range of ``healthy'' microbiomes in a given population.
For example, clinicians may pick a ``median'' health donor (i.e. one who is about as similar to all reference microbiomes), define a ``healthy plane'' and pick donors based on their distance to this plane (as in \cite{Halfvarson2017}), or simply based on the presence or abundance of certain consistently ``core'' health-associated bacteria \cite{Shade2011},\cite{Duvallet2017}.
In a similar vein, ``healthy'' donors can also be chosen based on their distance from disease-associated microbiomes, as opposed or in addition to similarity to health. Published case-control datasets can be used to identify donors with communities which are farthest away from the median or average diseased patient.
These datasets can also be mined to identify taxa which are consistently disease-associated, and which should be minimized or perhaps even absent in the potential donor.
Pairing rational donor selection with adaptive trial designs may eventually yield insight into the underlying model mediating the disease of interest if certain types of ``healthy'' donors consistently perform better at treating patients than others.

@@

The framework presented here encourages clinicians to leverage their clinical experience, existing microbiome research and published datasets, and the increasing availability of screened donor stools to more efficiently translate microbiome-based interventions into clinical impact.
Clinicians can apply their existing knowledge and a priori hypotheses to determine which microbiome-mediated disease model may underlie their indication of interest, and then select donors accordingly.
By rationally choosing donors during the FMT trial design, clinicians will increase the likelihood of successful FMT trials in diseases in which donor heterogeneity affects patient response.
Our power simulation analysis also suggests that specific plans for retrospective analyses of the microbiome data generated should be developed during trial design, with targeted hypotheses of interest and sample collection plans tailored accordingly.
Otherwise, exploratory analyses are unlikely to make new discoveries from most FMT trials. Paired with adaptive clinical trial designs, FMT trials with rationally-selected donors will become an important tool in advancing translational microbiome research and clinical treatment to improve and save patient lives.
